---
title: "team"
author: "Matthew Eisner"
date: "10/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(knitr)
```


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
defense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2019.xlsx', sheet='Biography')
```

## Part one: Cleaning data

Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset:

1. The player names and opponent names are factors.

```{r}
offense$Name <- as.factor(offense$Name)
offense$Opponent_Opponent <- as.factor(offense$Opponent_Opponent)

defense$Name <- as.factor(defense$Name)
defense$Opponent_Opponent <- as.factor(defense$Opponent_Opponent)

biography$Name <- as.factor(biography$Name)
```

2. All offensive and defensive statistics are numerical. The Weight column in biography is also numeric. (Optional: Learn to mutate multiple columns systematically using the `?across` function in `dplyr)

```{r}
offClean <- offense %>% mutate(across(where(is.character), as.numeric))
defClean <- defense %>% mutate(across(where(is.character), as.numeric))
bioClean <- biography %>% mutate(Weight = as.numeric(Weight))
```

3. Change the `Height` column in biography into numeric. (Hint: Use a unit that works well)

```{r}
bioClean <- bioClean %>%
  separate(Height, c('feet', 'inch'), sep='-') %>%
  mutate(feet = as.numeric(feet), inch = as.numeric(inch)) %>%
  mutate(Height_ft = feet + (inch/12)) %>%
  select(Name, Position, Height_ft, Weight, Class, Hometown, Highschool)
```

By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: `defClean`, `offClean`, and `bioClean`. Print the structures of these data frames.

```{r}
str(defClean)
str(offClean)
str(bioClean)
```

## Part two: Tidying

For the following questions, work on the cleaned data frames `defClean`, `offClean`, and `bioClean`.

1. Reformat the `defClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named `stat`.

```{r}

```

2. Compare the distributions of the defensive statistics. What defensive skills are rare?

3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.

Now turn our attention to the player biography.

4. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the `sep=` argument of `separate`)

5. How many players are there from each state?

## Part three: Joining data frames

1. Find from which states do the ISU players come. Create an appropriate summary.

2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.
