---
title: "team"
author: "Matthew Eisner"
date: "10/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(knitr)
```


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
defense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2019.xlsx', sheet='Biography')
```

## Part one: Cleaning data

Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset:

1. The player names and opponent names are factors.

```{r}
offense$Name <- as.factor(offense$Name)
offense$Opponent_Opponent <- as.factor(offense$Opponent_Opponent)

defense$Name <- as.factor(defense$Name)
defense$Opponent_Opponent <- as.factor(defense$Opponent_Opponent)

biography$Name <- as.factor(biography$Name)
```

2. All offensive and defensive statistics are numerical. The Weight column in biography is also numeric. (Optional: Learn to mutate multiple columns systematically using the `?across` function in `dplyr)

```{r}
offClean <- offense %>% mutate(across(where(is.character), as.numeric))
defClean <- defense %>% mutate(across(where(is.character), as.numeric))
bioClean <- biography %>% mutate(Weight = as.numeric(Weight))
```

3. Change the `Height` column in biography into numeric. (Hint: Use a unit that works well)

```{r}
bioClean <- bioClean %>%
  separate(Height, c('feet', 'inch'), sep='-') %>%
  mutate(feet = as.numeric(feet), inch = as.numeric(inch)) %>%
  mutate(Height_ft = feet + (inch/12)) %>%
  select(Name, Position, Height_ft, Weight, Class, Hometown, Highschool)
```

By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: `defClean`, `offClean`, and `bioClean`. Print the structures of these data frames.

```{r}
str(defClean)
str(offClean)
str(bioClean)
```

## Part two: Tidying

For the following questions, work on the cleaned data frames `defClean`, `offClean`, and `bioClean`.

1. Reformat the `defClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named `stat`.

```{r}
defClean <- defClean %>% pivot_longer(Tackles_Solo:Pass_PB, names_to='stat', values_to='count')
defClean
```

2. Compare the distributions of the defensive statistics. What defensive skills are rare?

```{r}
defClean %>% ggplot(aes(x=count)) + geom_histogram() + facet_wrap(~stat)
```

3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.

```{r}
dat <- defClean %>% 
  filter(stat=='Tackles_Solo' & Opponent_Opponent %in% c('Notre Dame', 'Iowa')) %>%
  pivot_wider(names_from = Opponent_Opponent, values_from = count)

dat %>% ggplot(aes(x=Iowa, y=`Notre Dame`)) + geom_point()
```

Now turn our attention to the player biography.

4. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the `sep=` argument of `separate`)

```{r}
bioClean <- bioClean %>% separate(Hometown, c('City', 'State'), sep=', ')
head(bioClean)
```

5. How many players are there from each state?

```{r}
bioClean %>%
  group_by(State) %>%
  summarize(Number.of.Players=n())
```

## Part three: Joining data frames

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.

```{r}
offClean %>% 
  left_join(bioClean, by='Name') %>%
  group_by(Name, State) %>%
  summarize()

defClean %>% 
  left_join(bioClean, by='Name') %>%
  group_by(Name, State) %>%
  summarize()
```

2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

```{r fig.width=7, fig.height=10}
fantasy <- offClean %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
  mutate(fantasy_points = Rushing_YDS*0.1 + Rushing_TD*6 + 
           Receiving_REC + Receiving_YDS*0.1 + Receiving_TD*6 + 
           Passing_YDS*0.04 + Passing_TD*4 - Passing_INT*2)

fantasy %>%  
  group_by(Name, Opponent_Opponent, fantasy_points) %>%
  summarize()

fantasy %>% 
  filter(Opponent_Opponent %in% c('ULM','Notre Dame', 'Texas', 'Oklahoma State')) %>%
  filter(Name %in% c('Purdy, Brock', "Pettway, La'Michael", 'Hall, Breece', 'Kolar, Charlie', 'Jones, Deshaunte',
                     'Milton, Tarique', 'Allen, Chase')) %>%
  ggplot(aes(x=Name, weight=fantasy_points, fill=Opponent_Opponent)) + 
  geom_bar() +
  scale_y_continuous(expand = c(0, 0.1)) +
  coord_flip() +
  facet_grid(rows = vars(Opponent_Opponent), scales = "free_y", switch = "y", space = "free_y") +
  labs(
    title = "Fantasy Football points for ISU",
    subtitle = "2019 Football Season",
    y = "Fantasy Points") +
  theme_minimal(base_family = "Roboto Condensed") +
  theme(
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
    plot.title = element_text(size = 15, face = "bold"),
    strip.text.y = element_text(angle = 270, face = "bold"),
    strip.placement = "outside",
    axis.title.x = element_text(margin = margin(t = 0.5, b = 0.5, unit = "cm")),
    axis.title.y = element_blank(),
    axis.text = element_text(size = 10),
    legend.position = "none",
    panel.grid.major.y = element_blank())
```

3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.

```{r}
defense2018 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Defensive')
offense2018 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Offensive')

offense2018$Name <- as.factor(offense2018$Name)
offense2018$Opponent_Opponent <- as.factor(offense2018$Opponent_Opponent)
defense2018$Name <- as.factor(defense2018$Name)
defense2018$Opponent_Opponent <- as.factor(defense2018$Opponent_Opponent)
offense2018 <- offense2018 %>% mutate(across(where(is.character), as.numeric))
defense2018 <- defense2018 %>% mutate(across(where(is.character), as.numeric))

fantasy2018 <- offense2018 %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
  mutate(fantasy_points_2018 = Rushing_YDS*0.1 + Rushing_TD*6 + 
           Receiving_REC + Receiving_YDS*0.1 + Receiving_TD*6 + 
           Passing_YDS*0.04 + Passing_TD*4 - Passing_INT*2)

fantasy %>% 
  select(-Rushing_YDS, -Rushing_TD, -Receiving_REC, -Receiving_YDS, 
         -Receiving_TD, -Passing_YDS, -Passing_TD, -Passing_INT, 
         -Rushing_ATT, -`Passing_CMP-ATT`) %>%
  inner_join(fantasy2018 %>% select(Name, Opponent_Opponent, fantasy_points_2018), 
             by=c('Name', 'Opponent_Opponent')) %>%
  group_by(Name) %>%
  summarize(mean.points.2018=mean(fantasy_points_2018), mean.points.2019=mean(fantasy_points))
```

